The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-10-29 submission/manuscript.Rmd
intro and abstract edits

The division labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the development of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain. By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline identity and meiosis, such as _Dazl_ and _Stra8_. KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing. KDM5C represses germline genes during ESC to EpiLC differentiation, during which it is recruited to a subset of CGI-containing germline identity gene promoters to facilitate CpG methylation. However the majority of germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.

**notes**
- How does sex and development influence germline gene repression?
- How does KDM5C interact with known silencing machinery? 
- Do all germline genes use teh same repressive mechanisms

  aberrantly expressed , including _Dazl_ and _Stra8_.
  - Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.

- Distinguishing the germline from the soma is a key step in the evolution of multicellularity and sexual reproduction. 
- Germline gene repression is orchestrated by chromatin regulators and transcription factors. 
- Much of these discoveries have been made looking at the repression of key marker genes for germ cell specification in the early embryo,

- In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
- Germline genes have not been comprehensively assessed as a whole

@@

Separation between germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in the later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e. sex and tissue environment) influences the manifestation of germline gene misexpression. Intriguingly, impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, elucidating how cell context contributes to germline gene dysregulation will reveal novel mechanisms governing these pathologies.

Here, we employed genome-wide analyses to explore loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C). KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some testis germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, what mechanisms underlie their dysregulation, and how KDM5C interacts with other known germline silencing mechanisms.

>>>>

The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity. In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI). Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders. However, it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation. Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain. By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline fate, such as _Dazl_ and _Stra8_. KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing. KDM5C represses germline genes during ESC to EpiLC differentiation and is recruited to a subset of CGI-containing germline gene promoters to facilitate DNA CpG methylation. However, the majority of mouse germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells through ectopic activation by RFX transcription factors. Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.

@@

Separation between germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes employ the same silencing mechanisms, such as those involved in the later stages of oogenesis and spermatogenesis. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e. sex and tissue environment) influences the manifestation of germline gene misexpression. Intriguingly, impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, elucidating how cell context contributes to germline gene dysregulation will reveal novel mechanisms governing these pathologies.

Here, we employed genome-wide analyses to explore the loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C). KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while pathogenic germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some testis germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, what mechanisms underlie their dysregulation, and how KDM5C interacts with other known germline silencing mechanisms.
